```{r load2, echo=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(scales)
load("Data/airlinesat.rdata")
```

# Examining and Summarizing Data

Sources for this chapter:

* ggplot2: [https://ggplot2.tidyverse.org/](https://ggplot2.tidyverse.org/)

## Introduction

Examining and summarizing data involves visualizations (e.g., graphs and charts) and tables. For visualization, the most popular package in R is th `ggplot2` package. For this part of the tutorial, the [airlinesat.rdata](Data/airlinesat.rdata) data file is used, which contains the <span style="font-family: Courier;">airlinesat</span> dataframe.  You should load this data now:<br>`load("airlinesat.rdata")`

\newpage
## Visualizations

### Package *ggplot2*

#### Introduction

* First, make sure the `ggplot2` library is loaded:<br>
`library(ggplot2)`
* `ggplot2` is based on the basic idea that for a graph, we need three things:
    1. Data: `ggplot2` is designed to work with data frames
    2. Coordinate system: what are the <span style="font-family: Courier;">x</span> and <span style="font-family: Courier;">y</span> variables
    3. Geoms: how is the data being represented (e.g., points, lines, etc.)  

    **AND** that the plot can be enhance by adding more layers, using `+`
* When ggplot is used in the console or from a script, the plot appears in the `Viewer` tab of the lower-right corner

#### Usage

* A plot starts with the `ggplot()` function, which requires two arguments:
    1. Source of the data, which can be piped in (i.e., `%>%`)
    2. The mapping of the data components
        * This argument is an aesthetic function, `aes()`, which maps the variable(s) to the coordinate system

    If the `ggplot()` function alone is used, the output is simply the coordinate system, but with nothing plotted
    * Because a geom hasn't been requested
    ```{r t2ggplotonly, fig.cap="Call to ggplot() without a geom", cache=TRUE}
    ggplot(airlinesat,    # Use data frame 'airlinesat'
           aes(x=country, y=nflights))   # Map 'country' on x and 'nflights' on y
    ```
* Other parts of the plot are adding in layers, using `+`
    * A good analogy is building a house:<br>
    The call to `ggplot()` is the foundation, but the structure is built one layer at a time
    * Example: Request a column chart for a discrete <span style="font-family: Courier;">x</span> and a continuous <span style="font-family: Courier;">y</span>
    ```{r t2simplecolumn, fig.cap="ggplot() + geom_col()", cache=TRUE}
    ggplot(airlinesat,    # Use data frame 'airlinesat'
           aes(x=country, y=nflights)) +  # Map 'country' on x and 'nflights' on y
       geom_col()   # Ask for column chart as the geom
    ```
    * **NOTE**: Each geom has a default statistic to plot
        * In this case, it is summing the `nflights` variable by country
        * We can use `dplyr` and `ggplot2` together to get a different value, such as the mean
    ```{r t2dplyrggplot, fig.cap="Using dplyr before ggplot to get mean values", cache=TRUE}
    # Use airlinesat data
    airlinesat %>%    
       # Group data by 'country'
       group_by(country) %>%
       # Create summary statistic
       summarise(mean_nflights=mean(nflights, na.rm=TRUE)) %>%
       # Pass this results to ggplot and start the plot
       ggplot(aes(x=country, y=mean_nflights)) +   # Note dataset was `piped`
         # Request column geom
         geom_col()
    ```

Using `ggplot()` can get much more advanced.  As the tutorial progresses, many examples of additional layers to a `ggplot()` will be shown.

### Bar and Column Charts

In `ggplot`, bar charts, `geom_bar()`, are used for plotting a single discrete variable, while column charts, `geom_col()`, are used for plotting a discrete variable on the <span style="font-family: Courier;">x</span> axis and a continuous variable on the <span style="font-family: Courier;">y</span> axis.

#### Bar Charts

* The standard bar chart provides a count of observations of each category of  discrete variable <span style="font-family: Courier;">x</span>
    ```{r t2bar1, fig.cap="Standard bar chart", cache=TRUE}
    airlinesat %>%
       ggplot(aes(x=gender)) +
         geom_bar()
    ```
* To get percentages of each category, we need to summarize the data and calculate the proportion for each category
    ```{r t2bar2, fig.cap="Bar chart with proportions", cache=TRUE}
    airlinesat %>%
       group_by(gender) %>%  # Group data by gender
       summarise(n=n()) %>%  # Create variable with count of each gender
       mutate(prop=n/sum(n)) %>%  # Create variable with proportion by gender
       ggplot(aes(x=gender,  # Variable for the x-axis
                  y=prop)) + # Use 'prop' instead of default counts for y-axis
         geom_bar(stat="identity")  # Use the value of y as-is
    ```
* To make the chart "pretty", we change the color of each bar we can add layers for axis labels, use the `scales` package to have the y-axis show percent, add labels for the bars etc.
    ```{r t2bar3, fig.cap="Prettier bar chart", cache=TRUE}
    airlinesat %>%
   group_by(gender) %>%  # Group data by gender
   summarise(n=n()) %>%  # Create variable with count of each gender
   mutate(prop=n/sum(n)) %>%  # Create variable with proportion by gender
   ggplot(aes(x=gender,  # Variable for the x-axis
              y=prop,   # Use 'prop' instead of default counts for y-axis
              fill=gender)) + # Use different color for each bar
      geom_bar(show.legend=FALSE, # Hide legend
               stat="identity", ) +  # Use the value of y as-is
      scale_y_continuous(labels=scales::label_percent()) + # y-axis labels %
      labs(x="Gender", y="Percent") +  # Label x- and y-axes
      geom_text(aes(label=sprintf("%.1f%%", prop*100)), # Format label number
             vjust=.95, # Vertically adjust the labels
             fontface="bold",  # Bold typeface
             color="white") # Text color
    ```

##### Bar Chart Variations

###### Stacked Bar Chart

Used to show one discrete variable by another discrete variable, such as data you would see in a cross-tab

* The <span style="font-family: Courier;">x=</span> variable specifies the axis, while the <span style="font-family: Courier;">fill=</span> variable stacks the bars by the other variable
* As with other bar charts, the default is to count observations, so some manipulation is needed to get "100% stacked bar charts"

```{r t2barchart3, fig.cap="Stacked bar chart", message=FALSE, cache=TRUE}
    airlinesat %>%
    group_by(gender, flight_type) %>%  # Group data by two discrete variables
    summarise(n=n()) %>%  # Count observations for each combination
    mutate(prop=n/sum(n)) %>%  # Calculate prop WITHIN first grouping variable
    ggplot(aes(x=gender, y=prop, fill=flight_type)) +
      geom_bar(position="fill",  # Stack the bars
               stat="identity") + # Use the value of y as-is
      scale_y_continuous(labels=scales::label_percent()) + # y-axis labels %
      labs(x="Gender", y="Percent",  # Label x- and y-axes 
           fill="Flight Type") +  # Label legend
      geom_text(aes(label=sprintf("%.1f%%", prop*100)), # Format data label
             position=position_stack(vjust=.95), # Adjust the labels
             fontface="bold",  # Bold typeface
             color="white") # Text color
```

###### Side-by-Side Bar Chart

Also used to shows one discrete variable by another discrete variable

* Again, default is to count observations, so some manipulation required to get percentages
    * Percentages can be within a group (like in 100% stacked, see Figure \@ref(fig:t2barchart4)) or percent of overall total (see Figure \@ref(fig:t2barchart5))
    ```{r t2barchart4, message=FALSE, fig.cap="Side-by-side bar chart (% within group)", cache=TRUE}
    # NOTE: The code for this chart is nearly identical to the previous figure
    #       ONLY the changes have been commented on below
    airlinesat %>%
    group_by(gender, flight_type) %>%  
    summarise(n=n()) %>%  
     mutate(prop=n/sum(n)) %>%  
    ggplot(aes(x=gender, y=prop, fill=flight_type)) +
      # NOTE: Use position="dodge" to make bars side-by-side
      geom_bar(position="dodge",  
               stat="identity") + 
      scale_y_continuous(labels=scales::label_percent()) + 
      labs(x="Gender", y="Percent",   
           fill="Flight Type") +  
      # NOTE: Use position=position_dodge(width=1) to position labels
      #       in center of each bar horizontally; use vjust=.95 to
      #       position labels at the top of each bar
      geom_text(aes(label=sprintf("%.1f%%", prop*100)), 
                position=position_dodge(width=1), 
                vjust=.95,
                fontface="bold",
                color="white")
    ```
    ```{r t2barchart5, message=FALSE, fig.cap="Side-by-side bar chart (% of total)", cache=TRUE}
    # NOTE: The code for this chart is nearly identical to the previous figure
    #       ONLY the changes have been commented on below
    airlinesat %>%
    group_by(gender, flight_type) %>%
      # NOTE: Use .groups="drop" to remove the grouping structure after
      #       summarising the data
      summarise(n=n(), .groups="drop") %>%  
      mutate(prop=n/sum(n)) %>%  
      ggplot(aes(x=gender, y=prop, fill=flight_type)) +
        geom_bar(position="dodge",  
                 stat="identity") + 
        scale_y_continuous(labels=scales::label_percent()) + 
        labs(x="Gender", y="Percent",   
             fill="Flight Type") +  
        geom_text(aes(label=sprintf("%.1f%%", prop*100)), 
                  position=position_dodge(width=1), 
                  vjust=.95,
                  fontface="bold",
                  color="white")
    ```

#### Column Charts

* The standard column chart provides a sum of continuous variable <span style="font-family: Courier;">y</span> of each category of disrete variable <span style="font-family: Courier;">x</span>
    ``` {r t2colchart1, fig.cap="Standard column chart", cache=TRUE}
    airlinesat %>%
      ggplot(aes(x=flight_type, y=nflights)) +
        geom_col()
    ```
    
* To get a different summary statistic, such as mean, we can summarize the data and calculate the summary statistic for each category (and make the graph prettier)
    ```{r t2colchart2, message=FALSE, fig.cap="Column chart showing means", cache=TRUE}
    airlinesat %>%
      group_by(flight_type) %>%
      summarise(mean=mean(nflights)) %>%
      ggplot(aes(x=flight_type, y=mean, fill=flight_type)) + 
        geom_col(show.legend=FALSE) +
        labs(x="Flight Type", y="Mean Number of Flights") + 
        geom_text(aes(label=sprintf("%.2f", mean)), # Format label number
             vjust=.95, # Vertically adjust the labels
             fontface="bold",  # Bold typeface
             color="white") # Text color
    ```

##### Side-by-Side Column Chart

* A side by side column chart can be used to show two discrete variables on the <span style="font-family: Courier;">x</span>-axis
    ```{r t2colchart3, message=FALSE, fig.cap="Side-by-side column chart", cache=TRUE}
    airlinesat %>%
    group_by(flight_type, flight_purpose) %>%  
    summarise(mean=mean(nflights), .groups="drop") %>%
    ggplot(aes(x=flight_type, y=mean, fill=flight_purpose)) +
      geom_col(position="dodge") + 
      labs(x="Flight Type", y="Mean Number of Flights",   
           fill="Flight Purpose") +  
      geom_text(aes(label=sprintf("%.2f", mean)), 
                position=position_dodge(width=1), 
                vjust=.95,
                fontface="bold",
                color="white")
    ```

### Histogram

In `ggplot`, histograms are produced with the `geom_histogram()` geom, which produces a histogram of a single continuous variable.

* By default, the <span style="font-family: Courier;">y</span>-axis is a count of observations in each "bin" of the <span style="font-family: Courier;">x</span> variable
    * A bin is a range of values of the continuous <span style="font-family: Courier;">x</span> variable
    * By default, `ggplot` will produce a histogram with 30 bins, and a message is produced to that effect unless the bins are changed manually
    ```{r t2hist1, fig.cap="Standard histogram", cache=TRUE}
    airlinesat %>%
      ggplot(aes(x=age)) +
        geom_histogram()
    ```

#### Changing Bins

Histograms can look quite different based on the bins used.  Bins can be changed in two ways: (1) number of bins; and (2) bin width

* Changing the number of bins is done with the `bins=` option
    * For example: `geom_histogram(bins=20)`
* Changing the bin width is done with the `binwidth=` option
    * For example; `geom_histogram(binwidth=5)`
* Use the interactive histograms (Figure \@ref(fig:t2shiny1) and Figure \@ref(fig:t2shiny2) to see how the histograms change

``` {r t2shiny1, echo=FALSE, fig.cap="Interactive histogram for number of bins"}
knitr::include_app("https://jdmeyer73.shinyapps.io/Histogram/", height="650px")
```
``` {r t2shiny2, echo=FALSE, fig.cap="Interactive histogram for bin width"}
knitr::include_app("https://jdmeyer73.shinyapps.io/Histogram2/", height="650px")
```

#### Improving the Look

You may find the default histogram a little "blah" or tough to read.  Just as the look of bar and column charts could be changed, so can the look of histograms
``` {r t2hist2, fig.cap="Prettier histogram", message=FALSE, cache=TRUE}
    airlinesat %>%
      ggplot(aes(x=age)) +
        geom_histogram(color="black",  # Adds red border around each bar
                       fill="tan") +  # Makes each bar blue
        labs(x="Age", y="Frequency")
```

#### Other Options

* Instead of the default count of observations, a density histogram can be created, where the sum of the area of the bars adds up to 1
* Often, a normal curve is added

look of bar and column charts could be changed, so can the look of histograms
``` {r t2hist3, fig.cap="Density histogram with normal curve", message=FALSE, cache=TRUE}
    airlinesat %>%
      ggplot(aes(x=age)) +
        geom_histogram(aes(y=..density..), # Request density instead of count
                       color="black",  # Adds red border around each bar
                       fill="tan") +  # Makes each bar blue
        stat_function(fun=function(x)   # Adds normal curve ovarlay
          dnorm(x, 
                mean=mean(airlinesat$age, na.rm=TRUE), # Mean of normal dist
                sd=sd(airlinesat$age, na.rm=TRUE))) +  # StdDev of normal dist
        labs(x="Age", y="Density")
```

### Box Plot

Box Plots are drawn with the `geom_boxplot()` geom, which by default creates a box plot for a continuous <span style="font-family: Courier;">y</span> variable, but for each level of a discrete <span style="font-family: Courier;">x</span> variable. In addition, the standard box plot does not contain "whiskers".

* To get a box plot for only the continuous <span style="font-family: Courier;">y</span> variable, use `x=""` as the discrete <span style="font-family: Courier;">x</span> variable
* To add whiskers, include a `stat_boxplot(geom="errorbar")` layer
    ```{r t2boxplot1, fig.cap="Single box plot with whiskers", cache=TRUE}
    airlinesat %>%
      ggplot(aes(x="", y=age)) +
        geom_boxplot() +
        stat_boxplot(geom="errorbar") +  # Add whiskers to box plot
        labs(x="",  # Remove x axis label
             y="Age")   # Make y axis label nicer
    ```

* To make comparisons across a discrete <span style="font-family: Courier;">x</span> variable, replaces the `x=""` from before with `x=VARIABLE`
    ```{r t2boxplot2, fig.cap="Multiple box plot", cache=TRUE}
    airlinesat %>%
      ggplot(aes(x=flight_purpose, y=age)) +
        geom_boxplot() +
        stat_boxplot(geom="errorbar") +  # Add whiskers to box plot
        labs(x="Flight Purpose", y="Age")
    ```

### Scatterplot

* Scatterplots are drawn with the `geom_point()` geom and are used to show the relationship between two continuous variables
    * Notice the warning given due to missing values (these warnings will be suppressed in other scatterplots below)
    ```{r t2scatter1, fig.cap="Standard scatterplot", cache=TRUE}
    airlinesat %>%
      ggplot(aes(x=age, y=s10)) + # s10 is satisfaction with condition of airplane
        geom_point()
    ```

#### Trendline
* Scatterplots become more helpful when we add a trend line.
* The most common trend line is a simple regression line, although others can be used.
    * Use `geom_smooth(method="lm", se=FALSE)` to add a linear trend line
    ```{r t2scatter2, fig.cap="Scatterplot with trendline", warning=FALSE, message=FALSE, cache=TRUE}
    airlinesat %>%
      ggplot(aes(x=age, y=s10)) + # s10 is satisfaction with condition of airplane
        geom_point() + 
        geom_smooth(method="lm", se=FALSE) +  # Add trendline
        labs(x="Age", y="Satisfaction with Aircraft Condition")
    ```    

#### Other Options
* The color, [https://ggplot2.tidyverse.org/articles/ggplot2-specs.html#sec:shape-spec](shape), and size of the points can be changed
    * In addition, they can vary by levels of a discrete variable
        * If a trend line is requested, separate trend lines will be provided for each level of the discrete variable
    ```{r t2scatter3, fig.cap="Scatterplot with different colors for discrete variable", warning=FALSE, message=FALSE, cache=TRUE}
    airlinesat %>%
      ggplot(aes(x=age, y=s10, color=flight_type)) + 
        geom_point(shape=17) + 
        geom_smooth(method="lm", se=FALSE) +  # Add trendline
        labs(x="Age", y="Satisfaction with Aircraft Condition",
             color="Flight Type")
    ```    

\newpage
## Tables and Statistics

### Frequency Table

* The `freq()` command from the package `questionr` produces nice one-way frequency tables (i.e., a frequency table for a single discrete variable)
    ```{r t2onewayfreq, cache=TRUE}
    library(questionr)
    freq(airlinesat$language, # Provide discrete variable
         cum=TRUE,  # Add cumulative percent column
         total=TRUE)  # Add total row at bottom
    ```
<caption>(\#tab:t2onewayfreq) One-way frequency table</caption> 
* Use package `htmlTable()` to produce a nicer looking table for output in the viewer (vs. the console)
    ```{r t2onewayfreq2, cache=TRUE}
    library(htmlTable)
    freq(airlinesat$flight_type, 
         cum=TRUE,  
         total=TRUE) %>%  # Pass result to htmlTable
      htmlTable()
    ```
<caption>(\#tab:t2onewayfreq2) One-way frequency table using <i>htmlTable</i></caption>

### Crosstabs

#### Base R

* Base R does not do a great job of easily creating cross-tabs and testing for independent of the two variables
* Using base R, a multistep process is required
    * Create the two-way frequency table using the `table(rowvar, colvar)` function and assign it to a separate object
        * Display the two-way freq table by just using the table name
    * Use the function `proportions(tablename, margin)` on the newly created object to get column, row, or total percentages
        * `proportions(tablename)` gives total percentages
        * `proportions(tablename, 1)` gives row percentages
        * `proportions(tablename, 2)` gives column percentages
    * Use the function `chisq.test(tablename)` on the newly created object to run the test of independence
    ```{r t2crosstab1, cache=TRUE}
    # Create two way table
    crosstab <- table(airlinesat$flight_purpose, # row Variable
                      airlinesat$gender)   # Column variable
    crosstab   # Display 2-way freq table
    proportions(crosstab, 2)  # Display column percentages
    chisq.test(crosstab)   # Run test of independence
    ```
<caption>(\#tab:t2crosstab1) Cross-tabs using Base R</caption>
* The look of the tables can be improved with the `htmlTable` package
    ```{r t2crosstab2, cache=TRUE}
    htmlTable(crosstab)
    htmlTable(round(proportions(crosstab,2)*100, 2))
    ```
<caption>(\#tab:t2crosstab2) Cross-tabs using <i>htmlTable</i></caption>

#### Alternative Packages

The following packages are not availabe through the BGSU Virtual Computing lab, but can be installed if using R/RStudio on your own machine.  These packages produce nicely formatted crosstabs.

##### Package *sjPlot*

* Use the function `tab_xtab(var.row=, var.col=, show.col.prc=TRUE)` to get a standard crosstab with column percentages
    ``` {r t2crosstab3, cache=TRUE, message=FALSE}
    library(sjPlot)
    tab_xtab(var.row=airlinesat$flight_purpose,
             var.col=airlinesat$gender,
             show.col.prc=TRUE)
    ```
<caption>(\#tab:t2crosstab3) Cross-tab using <i>sjPlot</i></caption>
    
##### Package *gmodels*

* Function`CrossTable(rowvar, colvar, OPTIONS)` has many options similar to SPSS
    ```{r t2crosstab4, cache=TRUE}
    library(gmodels)
    CrossTable(airlinesat$flight_purpose, airlinesat$gender,
               prop.r=FALSE,  # Exclude row percentages
               prop.t=FALSE,  # Exclude total percentages,
               prop.chisq=FALSE, # Exclude cell contribution to chi-sq
               digits=2,  # 2 digits after decimal point
               chisq=TRUE,  # Request test of independence
               format="SPSS")  # Request SPSS formatting
    ```
<caption>(\#tab:t2crosstab4) Cross-tab using <i>gmodels</i></caption>
    
### Measures of Centrality and Dispersion

#### Base R

* Any individual summary statistic can be easily calculated using Base R with functions such as:
    * `mean(x)` for mean
    * `sd(x)` for standard deviation
    * `quantile(x, .percentile)` for percentiles (e.g., '.50' would be median)
* For summary statistics except for standard deviation, the `summary(object)` function can be used, where object can be a single variable or an entire data frame
    ```{r t2summary1, cache=TRUE}
    # Summary for a single variable
    summary(airlinesat$nflights)
    ```
<caption>(\#tab:t2summary1) Summary statistics in R Base, one variable</caption>
    ```{r t2summary2, cache=TRUE}
    # Subset of airlinesat
    summary(airlinesat[,c("age", "nflights", "s10")])
    ```
<caption>(\#tab:t2summary2) Summary statistics in R Base, multiple variables</caption>
* Summary statistics for a continuous variable by different levels of a discrete variable can also be done in Base R using the `tapply(continuous variable, discrete variable, function)` function
    ``` {r t2summary3, cache=TRUE}
    tapply(airlinesat$nflights,  # Continous variable to apply the function to
           airlinesat$flight_purpose, # Discrete, grouping variable
           summary)  # R function to apply by group
    ```
<caption>(\#tab:t2summary3) Summary statistics in R Base, one variable, grouped</caption>

#### Package *dplyr*

* The `dplyr` package can also be used to manually create tables of summary statistics
    * One continuous variable
    ```{r t2summary4, cache=TRUE}
    airlinesat %>%
      summarise(mean=mean(age),
                sd=sd(age),
                q1=quantile(age, .25),
                median=quantile(age,.50),
                q3=quantile(age, .75))
    ```
<caption>(\#tab:t2summary4) Summary statistics using <i>dplyr</i>, one variable</caption>
    * Multiple continuous variables
    ```{r t2summary5, cache=TRUE}
    airlinesat %>%
      select(age, nflights, s10) %>%
      summary()
    ```
<caption>(\#tab:t2summary5) Summary statistics using <i>dplyr</i>, multiple variables</caption>
    * One continuous variable by a discrete/grouping variable
    ```{r t2summary6, cache=TRUE, tab.cap="Table Caption"}
    airlinesat %>%
      group_by(flight_purpose) %>%
      summarise(mean=mean(age),
                sd=sd(age),
                q1=quantile(age, .25),
                median=quantile(age,.50),
                q3=quantile(age, .75))
    ```
<caption>(\#tab:t2summary6) Summary statistics using <i>dplyr</i>, one variable, grouped</caption>

#### Package *vtable*

* Package `vtable` produces very nice looking tables of summary statistics, but it isn't available in BGSU's Virtual Computer Lab.
* Use function `sumtable(data, vars="varname")` to produce the table
    * One continuous variable
    ``` {r t2summary7code, eval=FALSE, cache=TRUE, message=FALSE}
    library(vtable)
    sumtable(airlinesat, vars="nflights")
    ```
    ``` {r t2summary7, echo=FALSE, cache=TRUE, message=FALSE}
    library(vtable)
    sumtable(airlinesat, vars="nflights", out="return") %>%
      kable(caption=NULL)
    ```
<caption>(\#tab:t2summary7) Summary statistics using <i>vtable</i>, one variable</caption>
    * Multiple continuous variables
    ``` {r t2summary8code, eval=FALSE, cache=TRUE}
    sumtable(airlinesat, 
             vars=c("nflights","age","s10"), # Use `c()` for multiple variables
             add.median=TRUE)  # Request median
    ```
    ``` {r t2summary8, echo=FALSE, cache=TRUE}
    sumtable(airlinesat, 
             vars=c("nflights","age","s10"), # Use `c()` for multiple variables
             out="return", 
             add.median=TRUE) %>%  # Request median
      kable(caption=NULL)
    ```
<caption>(\#tab:t2summary8) Summary statistics using <i>vtable</i>, multiple variable</caption>
    * One or more continuous variables by a discrete/grouping variable
    ``` {r t2summary9code, eval=FALSE, cache=TRUE}
    sumtable(airlinesat, 
             vars=c("nflights","age","s10"), 
             add.median=TRUE, 
             group="flight_purpose")
    ```
    ``` {r t2summary9, echo=FALSE, cache=TRUE}
    sumtable(airlinesat, 
             vars=c("nflights","age","s10"), 
             add.median=TRUE, 
             group="flight_purpose", out="return") %>%
      kable(caption=NULL)
    ```
<caption>(\#tab:t2summary9) Summary statistics using <i>vtable</i>, multiple variables, grouped</caption>

### Correlation

Correlation provides a measure of the strength of association between two continuous variables.

#### Base R

* Base R can easily provide the correlation and a test of the correlation using the `cor.test(variable1, variable2)` function
    * By default, it includes only observations that are non-missing in both variables
    ```{r t2corr1, cache=TRUE}
    cor.test(airlinesat$age, airlinesat$nflights)
    ```
<caption>(\#tab:t2corr1) Correlation with test in Base R</caption>
* Base R can also easily provide a correlation matrix of variables using the `cor(data)` function
    * By default, correlation will only be calculated for those pairs of variables that have no missing values
    * Use option `use="pairwise.complete.obs"` to exclude observations that are non-missing in both variables
    * However, Base R cannot produce a correlation matrix with p-values
    ```{r t2corr2, cache=TRUE}
    # First create data frame with only variables wanted
    mycorr <- airlinesat[,c("age", "nflights", "s10")]
    # Use function `round` to limit to 3 digits after decimal point
    round(cor(mycorr, use="pairwise.complete.obs"), 3)
    ```
<caption>(\#tab:t2corr2) Correlation matrix in Base R</caption>

#### Package *Hmisc*

* The function `rcorr()` from the package `Hmisc`, which is available in the BGSU Virtual Computing Lab, can be used to create correlation matrices also
    * The `rcorr()` function requires a matrix, so the data frame of variables must be coerced into a matrix first
    * By default, `rcorr()` produces three separate matrices: correlation, number of observations, and p-values
    * Separate tables can be requested
        * `rcorr(as.matrix(dataframe))]]"r"]]` provides the correlation matrix
        * `rcorr(as.matrix(dataframe))]]"P"]]` provides the matrix of p-values
        * Separate tables can used with `htmlTable()` for nicer output
    ``` {r t2corr3, cache=TRUE, message=FALSE}
    library(Hmisc)
    rcorr(as.matrix(mycorr))  # NOTE: 'mycorr' created in previous code
    ```
<caption>(\#tab:t2corr3) Correlation matrix using <i>Hmisc</i></caption>
    ```{r t2corr4code, eval=FALSE, cache=TRUE}
    # Use 'round()' function to limit digits in output
    htmlTable(round(rcorr(as.matrix(mycorr))[["r"]],4))
    htmlTable(round(rcorr(as.matrix(mycorr))[["P"]],5))
    ```
    ```{r t2corr41, echo=FALSE, cache=TRUE}
    htmlTable(round(rcorr(as.matrix(mycorr))[["r"]],4))
    ```
    ```{r t2corr42, echo=FALSE, cache=TRUE}
    htmlTable(round(rcorr(as.matrix(mycorr))[["P"]],5))
    ```
<caption>(\#tab:t2corr42) Separate correlation matrix output using <i>Hmisc</i></caption>

#### Package *sjPlot*

* The function `tab_corr()` from the `sjPlot` package produces very nice correlation matrices
    * `sjPlot` is not available in BGSU's Virtual Computing Lab
    ```{r t2corr5, cache=TRUE, message=FALSE}
    library(sjPlot)
    tab_corr(mycorr,   # Data frame of variables to use; created earlier
             na.deletion = "pairwise",  # Delete obs if either variable is missing
             corr.method = "pearson",  # Choose Pearson correlation coefficient
             show.p = TRUE,  # Show asterisks for significant correlations
             digits = 3,  # Show three decimal points
             triangle = "lower",  # Show only lower triangle
             fade.ns=FALSE)  # Do not fade insignficant correlations)
    ```
<caption>(\#tab:t2corr5) Correlation matrix output using <i>sjPlot</i></caption>

#### Package *GGally*

* The `ggpairs()` function from package `GGally` can produce a combination scatterplot and correlation matrix
    ```{r t2corr6, fig.cap="Combination scatterplot/correlation matrix using <i>GGally</i>", message=FALSE, warning=FALSE}
    library(GGally)
    ggpairs(mycorr, # Data frame created earlier
            lower=list(continuous=wrap("smooth",    # Adds fit lines...
                                       method="lm", # Using linear regression...
                                       se=FALSE,    # Without CI bands
                                       color="blue")),  # Color dots
            diag=list(continuous="blankDiag"))      # Sets diagonals to be blank
    ```
    