```{r echo=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(cowplot)
```
# Linear Regression

Sources for this chapter:

* *R for Marketing Research ad Analytics, Second Edition* (2019). Chris Chapman and Elea McDonnell Feit
    * [http://r-marketing.r-forge.r-project.org/index.html](http://r-marketing.r-forge.r-project.org/index.html)
* ggplot2: [https://ggplot2.tidyverse.org/](https://ggplot2.tidyverse.org/)

Data for this chapter:

* Once again, the [airlinesat.rdata](Data/airlinesat.rdata) is used. Load it now.
    ```{r t3loaddata}
    # You may need to change the directory
    load("Data/airlinesat.rdata")
    ```

## Introduction

Base R is typically sufficient for performing most regression tasks.  Some additional packages may be used for "prettier" tables or extracting results to make useful plots.

\newpage
## The `lm()` Function

* Basic linear regression is performed using the `lm()` function.
* Usage: `lm(formula, data)`
* In R, a <span style="font-family: Courier;">formula</span> is represented by dependent variables on the left side separated from the independent variables on the right side by a tilde(`~`), such as: `dv ~ iv1 + iv2`
    * For interactions between independent variables, use either `*` or `:`
        * `*` will include the interaction term AND each main effect
        * `:` will include ONLY the iteraction term
        * Examples:
            * `y ~ x1 + x2*x3` is the same as:<br>
            $y=x_1+x_2+x_3+(x_2\times x_3)$
            * `y ~ x1 + x2:x3` is the same as:<br>
            $y=x_1+(x_2\times x_3)$
            * `y ~ x1 + x2 + x2:x3` is the same as:<br>
            $y=x_1+x_2+(x_2\times x_3)$
* If `lm()` is run by itself, R only outputs the coefficients
    ```{r t3lm1}
    lm(nps ~ age + nflights, airlinesat)
    ```
<caption>(\#tab:t3lm1) Coefficients from `lm()` call</caption>

* However, if the results of the `lm()` call are assigned to an object, the `summary()` function can be used to get much more detailed output
    ```{r t3lm2, cache=TRUE}
    model1 <- lm(nps ~ age + nflights, airlinesat)
    summary(model1)
    ```
<caption>(\#tab:t3lm2) Summary results from `lm()` call</caption>
* For "nicer" looking results, the package `jtools` can be used
    * NOTE: `jtools` is no available in BGSU's Virtual Computing Lab
    ```{r t3lm3, cache=TRUE}
    library(jtools)
    summ(model1,   # Saved object from before
         digits=4, # How many digits to display in each column
         model.info = FALSE)  # Suppress extraneous information
    ```
<caption>(\#tab:t3lm3) Summary results from `lm()` using `jtools` package</caption>

* To get standardized beta coefficients, use the [lm_beta](./lm_beta.R) user defined function. 
    ```{r t3lmbeta}
    source("lm_beta.R")  # Download lm_beta.R and put in working directory
    lm_beta(model1, # Saved object with results
            digits=4)  # Number of digits to diplsy
    
    ```
<caption>(\#tab:t3lmbeta) Standardized Beta Coefficients</caption>

\newpage
## Prediction

The function `predict.lm()` can be used to predict the DV based on values of the IVs. This function is used in the margin plots covered in the next two sections. To use this function, we must pass a data frame of values to the function, where the data frame contains ALL of the IVs and the value for each IV that we want.

Suppose we wanted to predict, with a confidence interval, the `nps` of someone that is 45 years old and had 25 flights on the airline, and also someone that is 25 years old and had 45 flights on the airline. First, we create the data frame of values (see \@ref(using-vectors-in-line)):
```{r t3prediction1, cache=TRUE}
values <- data.frame(age=c(45, 25), nflights=c(25, 45))  # Create data frame
values  # Verify values
```
Second the data frame is passed to the `predict.lm()` function with confidence intervals requested:
```{r t3prediction2, cache=TRUE}
predict.lm(model1,   # The model we are using to predict
           values,   # The data frame of values to predict with
           interval="confidence")   # Request confidence interval
```

\newpage
## Margin Plots {#linregmarginplots}

In R, margin plots are a multistep process:

1. Use the `predictorEffect` function from the `effects` package to create a data frame of predicted values for different values of the focal variable.
    * For continuous focal variables, predictions will be made for 50 evenly-spaced values of the focal variable at mean values of the other variables.
        * If the model has an interaction between a continuous focal variable and a factor variable, predictions will be made for each level of the factor variable.
        * If the model has an interaction between a continuous focal variable and another continuous variable, predictions will be made for 5 evenly-spaced values of the other continuous variable.
    * For factor focal variables, predictions will be made for each level of the factor variable at mean values of the other variables.
        * If the model has an interaction between a factor focal variable and a continuous variable, predictions will be made for 5 evenly-spaced values of the the continuous variable.
        * If the model has an interaction between a factor focal variable and another factor variable, predictions will be made for every combination of the two factor variables.
    * By default, the dataframe will contain the predicted value (`fit`) and values for a $95\%$ confidence interval (`lower` and `upper`)
2. Use `ggplot` to create the margin plot

### Continuous IV (No Interaction)

``` {r t3marginplot1, message=FALSE, fig.cap="Margin plot (continuous IV)"}
library(effects)  # Load package 'effects'

# Want to predict 'nps' for different levels of 'age'
age.pred <- data.frame(predictorEffect("age",  # Focal variable
                                       model1))

# Use 'age.pred' for margin plot
age.pred %>%
   ggplot(aes(x=age,   # Age on x-axis
              y=fit)) +   # 'nps' prediction on y-axis
   geom_line(size=1) +   # Draw predicted line
   geom_ribbon(aes(ymin=lower,  # Draws the confidence interval bands
                   ymax=upper),
               alpha=0.2) + # Sets transparency level
   labs(x="Age", y="Linear Prediciton")
```

### Separate Margin Plots (No Interaction)

* If creating separate plots for each continuous IV, it is best to have the prediction (<span style="font-family: Courier;">y</span>) variable have the same scale
    * Doing so might take some trial and error after looking at graphs once
* Use the function `plot_grid()` from the `cowplot` package to make the plots appear in the same figure
    * Save each plot as an object, then pass the objects to `plot_grid()`
```{r t3marginplot2, message=FALSE, fig.cap="Separate margin plots (continuous factor IVs)"}
library(cowplot)
# Using dataframe from previous chunk, ask for min/max of the 95% CI adjust axis
min(age.pred$lower)
max(age.pred$upper)
# Save previous plot (for 'age') as plot 1
plot1 <- age.pred %>%
    ggplot(aes(x=age,y=fit)) +
        geom_line(size=1) + 
        geom_ribbon(aes(ymin=lower, ymax=upper), alpha=0.2) +
        labs(x="Age", y="Linear Prediciton") +
        scale_y_continuous(limits=c(5.25,9.25))  # Set scale limits for y-axis

# Create second plot for 'nflights' using procedure above
# Because of outliers in 'nflights', use the 'focal.levels=' option for more
#    realistic predictions
nflights.pred <- data.frame(predictorEffect("nflights",  # Focal variable
                                            model1, 
                                            focal.levels=seq(0,150,3)))
min(nflights.pred$lower)
max(nflights.pred$upper)
    
plot2 <- nflights.pred %>%
    ggplot(aes(x=nflights, y=fit)) +
        geom_line(size=1) +
        geom_ribbon(aes(ymin=lower,ymax=upper),alpha=0.2) +
        labs(x="Number of Flights", y="Linear Prediciton")  +
        scale_y_continuous(limits=c(5.25,9.25))  # Set scale limits for y-axis
    
# Use 'cowplot' function 'plot_grid' to combine plots
cowplot::plot_grid(plot1,plot2)
```
    
### Factor IV (No Interaction)

* Margin plots for a factor IV without an interaction consist of a point and $95\%$ confidence interval error bars
```{r t3marginplot2b1, cache=TRUE, message=FALSE}
# Run new model with flight_purpose included
model2 <- lm(nps ~ age + nflights + flight_purpose, airlinesat)
summary(model2)
```
<caption>(\#tab:t3marginplot2b1) Summary results of model with factor variable</caption>
```{r t3marginplot2b2, cache=TRUE, message=FALSE, fig.cap="Margin plot (factor IV)"}
# Create data frame with values to be plotted
fp.pred <- data.frame(predictorEffect("flight_purpose", model2))

# Create plot

fp.pred %>%
    ggplot(aes(x=flight_purpose, y=fit, 
               group=1)) +   # Need to draw line between points
        geom_point(size=4) + 
        geom_line(color="orange") + 
        geom_errorbar(aes(ymin=lower, ymax=upper), width=.5) +
        labs(x="Flight Purpose", y="Linear Prediction")
```

### Continuous IV (Interaction with Factor IV)

* Margin plots when for a continuous IV interaction with a factor IV are done in much the same way, but require an `aes(color=factor, fill=factor)` to produce different colored lines for each level of the factor
```{r t3marginplot3, cache=TRUE, message=FALSE}
# Run new model with 'age' and 'flight_purpose' interaction
model3 <- lm(nps ~ age*flight_purpose + nflights, airlinesat)
summary(model3)
```
<caption>(\#tab:t3marginplot3) Summary results of model with continuous IV and factor IV interaction</caption>
```{r t3marginplot4, cache=TRUE, message=FALSE, fig.cap="Margin plot (continuous IV with factor IV interaction)"}
# Create data frame with values to be plotted
age.pred <- data.frame(predictorEffect("age", model3))

# Create plot

age.pred %>%
   ggplot(aes(x=age, y=fit,
              color=flight_purpose, fill=flight_purpose)) +   # color based on factor
    geom_line(size=1) +   # Draw predicted line
    geom_ribbon(aes(ymin=lower,  # Draws the confidence interval bands
                   ymax=upper),
                alpha=0.2) + # Sets transparency level
    labs(x="Age", y="Linear Prediciton", color="Flight Purpose", fill="Flight Purpose") +
    theme(legend.position="bottom")
```

### Factor IV (Interaction with Continuous IV)

* Margin plots for a factor IV interaction with a continuous IV are done in much the same way, but require an `facet_wrap(~continuousIV)` to produce different different plots for evenly-spaced values of the continuous IV

```{r t3marginplot5, cache=TRUE, message=FALSE, fig.cap="Margin plot (factor IV with continuous IV interaction)"}
# Create data frame with values to be plotted
#   By default, 5 evenly-spaced values for 'age' will be created, but for use
#   with 'facet_wrap', set 'xlevels=4'
fp.pred <- data.frame(predictorEffect("flight_purpose", model3, xlevels=4))

# Create plot

fp.pred %>%
    ggplot(aes(x=flight_purpose, y=fit, group=1)) +
        geom_point(size=4) + 
        geom_line(color="orange") + 
        geom_errorbar(aes(ymin=lower, ymax=upper), width=.5) +
        facet_wrap(~age) +
        labs(x="Flight Purpose", y="Linear Prediction")
```

### Continuous IV (Interaction with Continuous IV)

```{r t3marginplot6, cache=TRUE, message=FALSE}
# Run new model with 'age' and 'overall_sat' interaction
model4 <- lm(nps ~ age*overall_sat + nflights, airlinesat)
summary(model4)
```
<caption>(\#tab:t3marginplot6) Summary results of model with continuous IV and continuous IV interaction</caption>
```{r t3marginplot7, cache=TRUE, message=FALSE, fig.cap="Margin plot (continuous IV with continuous IV interaction)"}
# Create data frame with values to be plotted
#   By default, 5 evenly-spaced values for 'overall_sat' will be created, but for 
#   easier interpretation, set 'xlevels=4'
age.pred <- data.frame(predictorEffect("age", model4, xlevels=4))

# Create plot

age.pred %>%
   ggplot(aes(x=age, y=fit,
              color=factor(overall_sat), fill=factor(overall_sat))) +   # color based on factor
    geom_line(size=1) +   # Draw predicted line
    geom_ribbon(aes(ymin=lower,  # Draws the confidence interval bands
                   ymax=upper),
                alpha=0.2) + # Sets transparency level
    labs(x="Age", y="Linear Prediciton", color="Overall Satisfaction", fill="Overall Satisfaction") +
    theme(legend.position="bottom")
```