```{r echo=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
```
# Principal Components Analysis

Sources for this chapter:

* *R for Marketing Research and Analytics, Second Edition* (2019). Chris Chapman and Elea McDonnell Feit
    * [http://r-marketing.r-forge.r-project.org/index.html](http://r-marketing.r-forge.r-project.org/index.html)

Data for this chapter:

* The `greekbrands` data is used from the `MKT4320BGSU` course package. Load the package and use the `data()` function to load the data.
    ```{r t6loaddata}
    # Load the course package
    library(MKT4320BGSU)
    # Load the data
    data(greekbrands)
    ```

## Introduction

Base R is typically sufficient for performing the basics of principal components analysis, but to get some of the outputs required more easily and more efficiently, I have created a user defined function, which is part of the `MKT4320BGSU` package.

* `pcaex` provides an eigenvalue table, scree plot, unrotated and rotated factor loading tables, and the principal components R object

## Base R

### `prcomp()` function

The `prcomp()` function performs PCA

* Usage: `prcomp(formula, data=, scale=TRUE, rank=)` where:
    * `formula` is a formula with no response variables, but rather only the numeric variables to be included if not all variables in `data` are to be included
        * No response variable means the formula is written as:<br>
        `~var1 + var2 + var3 + ... + var4`
    * `data=` is the name of the dataframe
    * `scale=TRUE` standardizes the variables before running the PCA
    * `rank=` is the number of components to retain; default is all
* When saved to an object, the following components are saved:
    * `$sdev` is the standard deviations of the principal components (i.e., the square roots of the eigenvalues)
    * `$rotation` is the unroated factor loadings
    * `$x` is the factor scores
* Example: perform PCA on $serious$, $fun$, $bargain$, $trendy$, and $value$ with only two components retained

    ```{r}
    pcaout <- prcomp(~serious+fun+bargain+trendy+value, # Variables to include
                     data=greekbrands,  # Data frame to use
                     scale=TRUE,   # Standardize the variables
                     rank=2)   # Retain only first two components
    ```

### Eigenvalue table

* To get the eigenvalue table, we must square the `$sdev` component of the PCA object

    ```{r}
    eigtable <- data.frame(components=seq(1:5), # column with component #'s
                           pcaout$sdev^2)  # Eigenvalues
    eigtable
    ```

### Unrotated loadings

* Unrotated factor loadings are in the `$rotation` component of the PCA object

    ``` {r}
    pcaout$rotation
    ```
### Rotated loadings

* Rotated loadings are not automatically created. Instead, we must use the `varimax(pcaobject$rotation)$loadings` command to obtain them. By default, only rotated loadings greater than 0.4 are shown.

    ``` {r}
    varimax(pcaout$rotation)$loadings
    ```

## User Defined Function

* The `pcaex` user defined function can produce the results with one or two passes of the function
    * Requires the following packages:
        * `ggplot2`
        * `dplyr`
* The results should be saved to an object
* Usage: `pcaex(data, group="", pref="", comp=)` where:
    * `data` is the PCA variable data
    * `group=""` is the name of the grouping variable
        * Can be excluded if no grouping variable
    * `pref=""` is the name of the preference variable
        * Can be excluded if no preference variable
    * `comp=` is the number of components to retain
        * Default is `NULL` if all components are wanted
* Objects returned: 
    * If `comp` is NOT provided:
        * Scree plot (`$plot`) 
        * Table of eigenvalues (`$table`) 
    * If `comp` is provided:
        * Table of eigenvalues (`$table`)
        * Unrotated factor loading table (`$unrotated`)
        * Rotated factor loading table (`$rotated`)
        * PCA object (`$pcaobj`)
* When a `group=` variable is provided, the PCA will be performed on an aggregated data frame (i.e., mean values by group)

### Preparation

* We need to pass a data frame to the user defined function containing the variables to be used (and maybe a preference and group/brand variable, see below)
    * In our class, the data set will often be used for creating a perceptual map, so we may also have a grouping variable (e.g., brand or product name) and a preference variable
    * Package `dplyr` is usually the best tool for this
* For this tutorial, we will perform a PCA using the `greekbrands` dataframe
    * We will use only the following attributes:
        * $perform$, $leader$, $fun$, $serious$, $bargain$, $value$
        * The data also has a preference variable, $pref$ and a group variable, $brand$, but we do not always want to use them.
    
```{r}
library(dplyr)
# Store variables selected to 'pcadata1' (WITHOUT group and pref variables)
pcadata1 <- greekbrands %>%   
    select(perform, leader, fun, serious, bargain, value)

# Store variables selected to 'pcadata2' (INCLUDES group and pref variables)
pcadata2 <- greekbrands %>%   
    select(perform, leader, fun, serious, bargain, value,
           pref, brand)
```

### Examples
    
#### WITHOUT `group=` or `pref=` options

* All components
```{r, messages=FALSE}
    gb1.all <- pcaex(pcadata1)    # PCA data created earlier
    # Do not include 'comp' to get all components
    
    # Call gb1.all$table to get eigenvalue table
    gb1.all$table
    # Call gb1.all$plot to get scree plot
    gb1.all$plot
```
    
* Two components
```{r, messages=FALSE}
    gb1.2comp <- pcaex(pcadata1,    # PCA data created earlier
                       comp=2)   # Request 2 components
    
    # Call gb1.2comp$unrotated to get unrotated factor loadings
    gb1.2comp$unrotated
    # Call gb1.2comp$rotated to get rotated factor loadings
    gb1.2comp$rotated
```

#### WITH `group=` or `pref=` options

* All components
```{r, messages=FALSE}
    gb2.all <- pcaex(pcadata2,    # PCA data created earlier
                     group="brand",   # Grouping variable
                     pref="pref")    # Preference variable
    # Do not include 'comp' to get all components
    # Call gb2.all$table to get eigenvalue table
    gb2.all$table
    # Call gb2.all$plot to get scree plot
    gb2.all$plot
```
    
* Two components
```{r}
    gb2.2comp <- pcaex(pcadata2,    # PCA data created earlier
                       group="brand",   # Grouping variable
                       pref="pref",    # Preference variable
                       comp=2)   # Request 2 components

    # Call gb2.2comp$unrotated to get unrotated factor loadings
    gb2.2comp$unrotated
    # Call gb2.2comp$rotated to get rotated factor loadings
    gb2.2comp$rotated
```
