```{r echo=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(MKT4320BGSU)
```

# PCA Perceptual Maps

Sources for this chapter:

* *R for Marketing Research and Analytics, Second Edition* (2019). Chris Chapman and Elea McDonnell Feit
    * [http://r-marketing.r-forge.r-project.org/index.html](http://r-marketing.r-forge.r-project.org/index.html)

Data for this chapter:

* The `greekbrands` data is used from the `MKT4320BGSU` course package. Load the package and use the `data()` function to load the data.
    ```{r t7loaddata}
    # Load the course package
    library(MKT4320BGSU)
    # Load the data
    data(greekbrands)
    ```

## Introduction

Base R can produce basic perceptual maps from a PCA object, but cannot create a joint-space map easily. For joint-space maps, better looking perceptual maps, and to make the process easier, I have created a user defined function, which is part of the `MKT4320BGSU` package.

* `percmap` returns a `ggplot` created perceptual (or joint-space) map.

## User Defined Function

### Preparation

* For the user defined function, we must pass the function a data frame with the variables to be used in creating the principal components, the grouping variable (i.e., brand/product name), and if applicable, the preference variable
    * The data frame does not need to be aggregated, but must contain ONLY those variables
    
```{r}
library(dplyr)
# For a perceptual map, store variables selected using dplyr to 'pmdata'
pmdata <- greekbrands %>% 
    select(perform, leader, fun, serious, bargain, value, brand)

# For a joint space map, store variable selected using dplyr to 'jsmdata'
jsmdata <- greekbrands %>%
    select(perform, leader, fun, serious, bargain, value, pref, brand)
```

### Using the function

* The `percmap` user defined function can easily product a perceptual map or joint-space map based on the first two principal components
    * Requires the following packages:
        * `ggplot2`
        * `dplyr`
* Usage: `percmap(data, group="", pref="")` where:
    * `data` is the PCA variable data
    * `group=""` is the name of the grouping variable
    * `pref=""` is the name of the preference variable
        * Can be excluded if no preference variable
* Returns:
    * A perceptual map if `pref` is not included
    * A joint space map if `pref` is included
* Examples:
    * Perceptual Map
    ```{r}
    percmap(pmdata, group="brand")
    ```
    * Joint Space Map
    ```{r}
    percmap(jsmdata, group="brand", pref="pref")
    ```
