```{r echo=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
```

# PCA Perceptual Maps

Sources for this chapter:

* *R for Marketing Research ad Analytics, Second Edition* (2019). Chris Chapman and Elea McDonnell Feit
    * [http://r-marketing.r-forge.r-project.org/index.html](http://r-marketing.r-forge.r-project.org/index.html)

Data for this chapter:

* The [greekbrands.rdata](Data/greekbrands.rdata) is used. Load it now.
    ```{r}
    # You may need to change the directory
    load("Data/greekbrands.rdata")
    ```

## Introduction

Base R can produce basic perceptual maps from a PCA object, but cannot create a joint-space map easily. For joint-space maps, better looking perceptual maps, and to make the process easier, user defined function has been created

* [percmap.R](percmap.R) returns a `ggplot` created percpetual (or joint-space) map.

Once saved in your working directory, it is good practice to source user defined functions.
``` {r}
source("percmap.R")
```

## Base R

### Preparation

* For PCA perceptual maps with a grouping variable (e.g., brand or product name), it is best to use a dataframe of aggregated means by the grouping variable.
* For this tutorial, we will create a PCA perceptual map using the `greekbrands` dataframe
    * We will use only the following attributes:
        * $perform$, $leader$, $fun$, $serious$, $bargain$, $value$
        * The grouping variable is $brand$
    
```{r}
library(dplyr)
# Store variables selected using dplyr to 'pmdata'
pmdata <- greekbrands %>% 
    select(perform, leader, fun, serious, bargain, value, brand) %>%
    group_by(brand) %>%
    summarise_all(mean)
```

### Perform PCA first

* To create the perceptual map, we must first run and store a PCA object on our data frame

    ```{r}
    pcaout <- prcomp(~perform+leader+fun+serious+bargain+value, 
                     data=pmdata, 
                     scale=TRUE,   
                     rank=2)  
    ```

### Create map using `biplot`

* The map is created using the `biplot` function with the PCA object passed to it along with group variable labels

``` {r}
biplot(pcaout,  # PCA object from above
       xlabs=pmdata$brand)   # Grouping variable labels
```

## User Defined Function

### Preparation

* For the user defined function, we must pass the function a data frame with the variables to be used in creating the principal components, the grouping variable (i.e., brand/product name), and if applicable, the preference variable
    * The data frame does not need to be aggregated, but must contain ONLY those variables
    
```{r}
library(dplyr)
# For a perceptual map, store variables selected using dplyr to 'pmdata'
pmdata <- greekbrands %>% 
    select(perform, leader, fun, serious, bargain, value, brand)

# For a joint space map, store variable selected using dplyr to 'jsmdata'
jsmdata <- greekbrands %>%
    select(perform, leader, fun, serious, bargain, value, pref, brand)
```

### Using the function

* The `percmap.R` user defined function can easily product a perceptual map or joint-space map based on the first two principal components
    * Requires the following packages:
        * `ggplot2`
        * `dplyr`
* Usage: `percmap(data, group="", pref="")` where:
    * `data` is the PCA variable data
    * `group=""` is the name of the grouping variable
    * `pref=""` is the name of the preference variable
        * Can be excluded if no preference variable
* Returns:
    * A perceptual map if `pref` is not included
    * A joint space map if `pref` is included
* Examples:
    * Perceptual Map
    ```{r}
    source("percmap.R")
    percmap(pmdata, group="brand")
    ```
    * Joint Space Map
    ```{r}
    percmap(jsmdata, group="brand", pref="pref")
    ```
