```{r echo=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(MKT4320BGSU)
```

# Standard MNL

Sources for this chapter:

* *R for Marketing Research and Analytics, Second Edition* (2019). Chris Chapman and Elea McDonnell Feit
    * [http://r-marketing.r-forge.r-project.org/index.html](http://r-marketing.r-forge.r-project.org/index.html)

Data for this chapter:

* The [bfast.rdata](Data/bfast.rdata) is used, which is part of the `MKT4320BGSU` package.
    ```{r}
    # You may need to change the directory
    data(bfast)
    ```

## Introduction

Base R is not good for standard multinomial logistic regression (MNL). The best package that I have found for standard MNL is `nnet` with its `multinom` function. Install that package `install.packages("nnet")` and load it.

``` {r}
library(nnet)
```

In addition, I have created some user defined functions, which are all in the `MKT4320BGSU` package.

* `stmnl` produces Odds Ratio coefficients table, overall model significance, and McFadden's Pseudo-$R^2$
* `stmnl_cm1` produces a Classification Matrix
* `stmnl_pp` produces average predicted probability tables and plots

## Data Preparation

As with binary logistic regression, we often us a training and holdout sample when using MNL. For standard MNL, the process is the same.

``` {r}
library(caret)
set.seed(4320)
inTrain <- createDataPartition(y=bfast$bfast, p=.75, list=FALSE)
train <- bfast[inTrain,]
test <- bfast[-inTrain,]
```

## Standard MNL using `multinom()`

* Standard MNL is performed using the `multinom()` function from the `nnet` package
* Usage: `multinom(formula, data)`
    * `formula` is represented by dependent variables on the left side separated from the independent variables on the right side by a tilde(`~`), such as: `dv ~ iv1 + iv2`
    * `data` is the name of the (usually) training data
* As with other analyses, we save the result of the model to an object
    * `summary()` provides standard coefficient estimates (i.e., **not** Odds Ratio estiamtes), but does not provide overall model fit values (i.e., overall model $p$-value or McFadden's Psuedo $R^2$) or p-values for each independent variable

``` {r}
library(nnet)
model <- multinom(bfast ~ gender + marital + lifestyle + age, data=train)
summary(model)
```

### `stmnl()` User Defined Function

* To get overall model fit values, as well as an Odds Ratio estimate table with p-values, the `stmnl()` user defined function can be used
* Usage: `stmnl(model)`
    * `model` is the name of the object with the saved model results
* NOTE: This function requires the `broom` package.  If you do not have it installed on your machine, you will need to install it.

```{r}
stmnl(model)
```

### Classification Matrix

* To get a classification matrix, the `stmnl_cm` user-defined function should be used
* Usage: `stmnl_cm(model, data)`
    * `model` is the name of the object with the saved model results
    * `data` is name of the data to create the classification model for (i.e., training or holdout data)
    
```{r}
stmnl_cm(model, train)
stmnl_cm(model, test)
```

### Average Predicted Probabilities

* Predicted probabilities can help interpret the effects of the independent variables on the choice dependent variable
* Use the `stmnl_pp` user-defined function for each IV to obtain these.
* Usage: `stmnl_pp(model, focal, xlab)`
    * `model` is the name of the object with the saved model results
    * `focal` is the name of the variable (in quotes) for which predicted probabilities are wanted
    * `xlab` is optional, but can be provided for a better x-axis label for the plot (e.g., "Income Category")  
* NOTE 1: For continuous focal variables, the table produced uses three levels to calculate predicted probabilities: $-1 SD$, $MEAN$, $+1 SD$   
* NOTE 2: This function requires the following packages:
    * `tidyr`
    * `effects`
    * `dplyr`
    * `ggplot2`

```{r}
stmnl_pp(model, "age", "Age in Years")
stmnl_pp(model, "gender", "Gender")
```