```{r echo=FALSE, message=FALSE, warning=FALSE}
library(fpp3)
library(slider)
library(purrr)
library(ggplot2)
library(dplyr)
```

# Forecasting

Data for this chapter:

* The [msales.rdata](Data/msales.rdata) and [qsales.rdata](Data/qsales.rdata) are used. Load them now.
```{r}
# You may need to change the directory
load("Data/msales.rdata")
load("Data/qsales.rdata")
```

## Introduction

While Base R does have some time-series forecasting capabilities, there are several packages that make working with the data and running the analysis easier. For our class, we will be mainly using the `fpp3` package inside of some user-defined functions. The user-defined functions are described briefly below, and then in detail in their own sections.

### User-Defined Functions for Forecasting I

* [naivefc.R](naivefc.R) analyzes the data using naive forecasting methods
* [smoothfc.R](smoothfc.R) analyzes the data using smoothing forecasting methods
* [linregfc.R](linregfc.R) analyzes the data using regression-based forecasting methods

### User-Defined Functions for Forecasting II

* [acplots.R](acplots.R) creates ACF and PACF plots needed for ARIMA forecasting
* [tswh.noise.R](tswh.noise.R) creates a cumulative periodogram and white noise test needed for ARIMA forecasting
* [autoarima.R](autoarima.R) runs an ARIMA model as specified by the user and an automatic search for the "best" model if requested

### User-Defined Functions for Both

* [tsplot.R](tsplot.R) produces a time series plot of the data
* [fccompare.R](fccompare.R) compares models from saved results after running the methods functions

### Packages

You must have the following packages installed to use all of the user-defined functions. If you do not have them installed, you should do so now.

* `fpp3`
* `slider`
* `purrr`
* `dplyr`
* `ggplot2`
* `ggfortify`
* `stringr`
* `cowplot`

```{r, eval=FALSE}
# Load necessary packages
library(fpp3)
library(slider)
library(purrr)
library(ggplot2)
library(dplyr)
library(ggfortify)
library(stringr)
library(cowplot)
```

## `tsplot` User Defined Function

* Usage: `tsplot(data, tvar="", obs="", datetype=c("ym", "yq", "yw"), h= )`
    * `data` is the name of the dataframe with the time variable and measure variable
    * `tvar` is the variable that identifies the time period (in quotes)
    * `obs` is the variable that identifies the measure (in quotes)
    * `datetype` can be one of three options:
        * `"ym"` if the time period is year-month
        * `"yq"` if the time period is year-quarter
        * `"yw"` if the time period is year-week
    * `h` is an integer indicating the number of holdout/forecast periods
* Output: A time series plot

``` {r}
source("tsplot.R")
tsplot(msales,  # Dataframe
       "t",     # Date variable
       "sales", # Measure variable
       "ym",    # Date type
       12)      # Holdout periods
tsplot(qsales, "t", "sales", "yq", 8)
```

## `naivefc` User Defined Function

* Usage: `naivefc(data, tvar="", obs="", datetype=c("ym", "yq", "yw"), h= )`
    * `data` is the name of the dataframe with the time variable and measure variable
    * `tvar` is the variable that identifies the time period (in quotes)
    * `obs` is the variable that identifies the measure (in quotes)
    * `datetype` can be one of three options:
        * `"ym"` if the time period is year-month
        * `"yq"` if the time period is year-quarter
        * `"yw"` if the time period is year-week
    * `h` is an integer indicating the number of holdout/forecast periods
* NOTE 1: The results of this function should be saved to an object.  When doing so, the following objects are returned:
    * `$plot` contains the naive methods plot
    * `$acc` contains the accuracy measures
    * `$fcresplot` contains a plot of the holdout period residuals
    * `$fcresid` is a dataframe of the holdout period residuals (seldom used separately)
* Note 2: The model names are:
    * **Naive** for the naive model
    * **Naive.Drift** for the naive model with drift
    * **Seas.Naive** for the seasonal naive model
    * **Seas.Naive.Drift** for the seasonal naive model with drift

``` {r}
source("naivefc.R")
naive <- naivefc(msales, "t", "sales", "ym", 12)
naive$plot
naive$acc
naive$fcresplot
```

## `smoothfc` User Defined Function

* Usage: `smoothfc(data, tvar="", obs="", datetype=c("ym", "yq", "yw"), h= )`
    * `data` is the name of the dataframe with the time variable and measure variable
    * `tvar` is the variable that identifies the time period (in quotes)
    * `obs` is the variable that identifies the measure (in quotes)
    * `datetype` can be one of three options:
        * `"ym"` if the time period is year-month
        * `"yq"` if the time period is year-quarter
        * `"yw"` if the time period is year-week
    * `h` is an integer indicating the number of holdout/forecast periods
* NOTE 1: The results of this function should be saved to an object.  When doing so, the following objects are returned:
    * `$plot` contains the smoothing methods plot
    * `$acc` contains the accuracy measures
    * `$fcresplot` contains a plot of the holdout period residuals
    * `$fcresid` is a dataframe of the holdout period residuals (seldom used separately)
* Note 2: The model names are:
    * **Mov.Ave** for the moving average model
    * **Exp.Smooth** for the exponential smoothing model
    * **H-W.Add** for the Holt-Winters Additive model
    * **H-W.Mult** for Hold-Winters Multiplicative model

``` {r}
source("smoothfc.R")
smooth <- smoothfc(msales, "t", "sales", "ym", 12)
smooth$plot
smooth$acc
smooth$fcresplot
```

## `linregfc` User Defined Function

* Usage: `linregfc(data, tvar="", obs="", datetype=c("ym", "yq", "yw"), h= )`
    * `data` is the name of the dataframe with the time variable and measure variable
    * `tvar` is the variable that identifies the time period (in quotes)
    * `obs` is the variable that identifies the measure (in quotes)
    * `datetype` can be one of three options:
        * `"ym"` if the time period is year-month
        * `"yq"` if the time period is year-quarter
        * `"yw"` if the time period is year-week
    * `h` is an integer indicating the number of holdout/forecast periods
* NOTE 1: The results of this function should be saved to an object.  When doing so, the following objects are returned:
    * `$plot` contains the regression methods plot
    * `$acc` contains the accuracy measures
    * `$fcresplot` contains a plot of the holdout period residuals
    * `$fcresid` is a dataframe of the holdout period residuals (seldom used separately)
* Note 2: The model names are:
    * **Lin.Reg.Trend** for the linear regression with trend model
    * **Line.Reg.Seas.Trend** for linear regression with trend and seasonality model
    
``` {r}
source("linregfc.R")
linreg <- linregfc(msales, "t", "sales", "ym", 12)
linreg$plot
linreg$acc
linreg$fcresplot
```

## `acplots` User Defined Function

* Usage: `acplots(data, tvar="", obs="", datetype=c("ym", "yq", "yw"), h= , lags=, d=, D=c(0,1))`
    * `data` is the name of the dataframe with the time variable and measure variable
    * `tvar` is the variable that identifies the time period (in quotes)
    * `obs` is the variable that identifies the measure (in quotes)
    * `datetype` can be one of three options:
        * `"ym"` if the time period is year-month
        * `"yq"` if the time period is year-quarter
        * `"yw"` if the time period is year-week
    * `h` is an integer indicating the number of holdout/forecast periods
    * `lags` is an integer indicating how many lags should be shown on the plot; default is 25
    * `d` is the order of non-seasonal differencing; default is 0
    * `D` is the order of seasonal differencing; default is 0
* NOTE: The results of this function can be saved to an object.  When doing so, the following objects are returned:
    * `$acf` contains the ACF plot
    * `$pacf` contains the PACF plot

``` {r}
source("acplots.R")
out <- acplots(msales, "t", "sales", "ym", 12,
               lags=25,   # Number of lags to plot
               d=0,       # Non-seasonal differencing
               D=1)       # Seasonal differencing
out$acf
out$pacf
```

## `tswh.noise` User Defined Function

* Usage: `tswh.noise(data, tvar="", obs="", datetype=c("ym", "yq", "yw"), h= , arima=c(p,d,q), sarima=c(P,D,Q))`
    * `data` is the name of the dataframe with the time variable and measure variable
    * `tvar` is the variable that identifies the time period (in quotes)
    * `obs` is the variable that identifies the measure (in quotes)
    * `datetype` can be one of three options:
        * `"ym"` if the time period is year-month
        * `"yq"` if the time period is year-quarter
        * `"yw"` if the time period is year-week
    * `h` is an integer indicating the number of holdout/forecast periods
    * `lags` is an integer indicating how many lags should be shown on the plot; default is 25
    * `arima=(p,d,q)` provides the non-seasonal autoregressive ($p$), differencing ($d$), and moving average ($q$) parameters, in the form of `c(p,d,q)` where the parameters are integers
    * `sarima=(P,D,Q)` provides the seasonal autoregressive ($P$), differencing ($D$), and moving average ($Q$) parameters, in the form of `c(P,D,Q)` where the parameters are integers
* The function will return a cumulative periodogram and results of the Ljung-Box White Noise test

``` {r}
source("tswh.noise.R")
tswh.noise(msales, "t", "sales", "ym", 12,
           arima=c(0,0,0),     # Seasonal parameters
           sarima=c(0,0,0))    # Non-seasonal parameters

tswh.noise(msales, "t", "sales", "ym", 12, c(2,1,0), c(2,1,0))

```

## `autoarima` User Defined Function

* Usage: `tswh.noise(data, tvar="", obs="", datetype=c("ym", "yq", "yw"), h= , arima=c(p,d,q), sarima=c(P,D,Q), auto="")`
    * `data` is the name of the dataframe with the time variable and measure variable
    * `tvar` is the variable that identifies the time period (in quotes)
    * `obs` is the variable that identifies the measure (in quotes)
    * `datetype` can be one of three options:
        * `"ym"` if the time period is year-month
        * `"yq"` if the time period is year-quarter
        * `"yw"` if the time period is year-week
    * `h` is an integer indicating the number of holdout/forecast periods
    * `lags` is an integer indicating how many lags should be shown on the plot; default is 25
    * `arima=(p,d,q)` provides the non-seasonal autoregressive ($p$), differencing ($d$), and moving average ($q$) parameters, in the form of `c(p,d,q)` where the parameters are integers
    * `sarima=(P,D,Q)` provides the seasonal autoregressive ($P$), differencing ($D$), and moving average ($Q$) parameters, in the form of `c(P,D,Q)` where the parameters are integers
    * `auto` can be one of two options:
        * `"Y"` an automatic search is performed to find the "best" model
        * `"N"` an automatic search is not performed
* NOTE 1: The results of this function should be saved to an object.  When doing so, the following objects are returned:
    * `$plot` contains the ARIMA methods plot
    * `$acc` contains the accuracy measures
    * `$fcresplot` contains a plot of the holdout period residuals
    * `$fcresid` is a dataframe of the holdout period residuals (seldom used separately)
    * `$acresid` contains a plot of the ACF/PACF residuals for the model(s)
    * `$wn` contains cumulative periodogram(s) and results of the Ljung-Box White Noise test(s)
* Note 2: If `auto="Y"` is specified, this function can take one to two minutes to run
* Note 3: The model names are:
    * **Self** for the model specified by the user
    * **Auto** for the "best" model if requested by the user

``` {r}
source("autoarima.R")
selfonly <- autoarima(msales, "t", "sales", "ym", 12,
                      arima=c(2,1,0),     # Seasonal parameters
                      sarima=c(2,1,0),    # Non-seasonal parameters
                      auto="N")           # Do not perform auto search
selfonly$plot
selfonly$acc
selfonly$fcresplot
selfonly$acresid
selfonly$wn
```

``` {r}
arima <- autoarima(msales, "t", "sales", "ym", 12,
                   arima=c(2,1,0),     # Seasonal parameters
                   sarima=c(2,1,0),    # Non-seasonal parameters
                   auto="Y")           # Do not perform auto search
arima$plot
arima$acc
arima$fcresplot
arima$acresid
arima$wn
```

## `fccompare` User Defined Function

* Usage: `fccompare(results, models)`
    * `results` is a list of the stored methods results; create the list with the `list` function, such as:<br>`results <- list(naive, smooth, linreg)`
    * `models` is a vector of the models you want to compare; create the vector with the `c()` function, such as:<br>`models <- c("Naive", "Mov.Ave", "Lin.Reg.Trend")`
* NOTE 1: The function returns two items that will be directly displayed or can be saved to an object:  
    * `$acc` contains the accuracy measures
    * `$fcresplot` contains a plot of the holdout period residuals

``` {r}
source("fccompare.R")
results <- list(naive, smooth, linreg, arima)
models <- c("Naive", "Mov.Ave", "Lin.Reg.Trend", "Auto")
fccompare(results, models)
```