```{r echo=FALSE, message=FALSE, warning=FALSE}
library(fpp3)
library(slider)
library(purrr)
library(ggplot2)
library(dplyr)
```

# Forecasting I

Data for this chapter:

* The `msales` and `qales` data are used from the `MKT4320BGSU` course package. Load the package and use the `data()` function to load the data.
    ```{r t11loaddata}
    # Load the course package
    library(MKT4320BGSU)
    # Load the data
    data(msales)
    data(qsales)
    ```

## Introduction

While Base R does have some time-series forecasting capabilities, there are several packages that make working with the data and running the analysis easier. For our class, we will be mainly using the `fpp3` package inside of some user-defined functions. The user-defined functions are described briefly below, and then in detail in their own sections.

### User-Defined Functions for Forecasting I

* `naivefc` analyzes the data using naive forecasting methods
* `smoothfc` analyzes the data using smoothing forecasting methods
* `linregfc` analyzes the data using regression-based forecasting methods
* `tsplot` produces a time series plot of the data
* `fccompare` compares models from saved results after running the methods functions

### Packages

You must have the following packages installed to use all of the user-defined functions. If you do not have them installed, you should do so now.

* `fpp3`
* `slider`
* `purrr`
* `dplyr`
* `ggplot2`
* `ggfortify`
* `stringr`
* `cowplot`

```{r, eval=FALSE}
# Load necessary packages
library(fpp3)
library(slider)
library(purrr)
library(ggplot2)
library(dplyr)
library(ggfortify)
library(stringr)
library(cowplot)
```

## `tsplot` User Defined Function

* Usage: `tsplot(data, tvar="", obs="", datetype=c("ym", "yq", "yw"), h= )`
    * `data` is the name of the dataframe with the time variable and measure variable
    * `tvar` is the variable that identifies the time period (in quotes)
    * `obs` is the variable that identifies the measure (in quotes)
    * `datetype` can be one of three options:
        * `"ym"` if the time period is year-month
        * `"yq"` if the time period is year-quarter
        * `"yw"` if the time period is year-week
    * `h` is an integer indicating the number of holdout/forecast periods
* Output: A time series plot

``` {r}
tsplot(msales,  # Dataframe
       "t",     # Date variable
       "sales", # Measure variable
       "ym",    # Date type
       12)      # Holdout periods
tsplot(qsales, "t", "sales", "yq", 8)
```

## `naivefc` User Defined Function

* Usage: `naivefc(data, tvar="", obs="", datetype=c("ym", "yq", "yw"), h= )`
    * `data` is the name of the dataframe with the time variable and measure variable
    * `tvar` is the variable that identifies the time period (in quotes)
    * `obs` is the variable that identifies the measure (in quotes)
    * `datetype` can be one of three options:
        * `"ym"` if the time period is year-month
        * `"yq"` if the time period is year-quarter
        * `"yw"` if the time period is year-week
    * `h` is an integer indicating the number of holdout/forecast periods
* NOTE 1: The results of this function should be saved to an object.  When doing so, the following objects are returned:
    * `$plot` contains the naive methods plot
    * `$acc` contains the accuracy measures
    * `$fcresplot` contains a plot of the holdout period residuals
    * `$fcresid` is a dataframe of the holdout period residuals (seldom used separately)
* Note 2: The model names are:
    * **Naive** for the naive model
    * **Naive.Drift** for the naive model with drift
    * **Seas.Naive** for the seasonal naive model
    * **Seas.Naive.Drift** for the seasonal naive model with drift

``` {r}
naive <- naivefc(msales, "t", "sales", "ym", 12)
naive$plot
naive$acc
naive$fcresplot
```

## `smoothfc` User Defined Function

* Usage: `smoothfc(data, tvar="", obs="", datetype=c("ym", "yq", "yw"), h= )`
    * `data` is the name of the dataframe with the time variable and measure variable
    * `tvar` is the variable that identifies the time period (in quotes)
    * `obs` is the variable that identifies the measure (in quotes)
    * `datetype` can be one of three options:
        * `"ym"` if the time period is year-month
        * `"yq"` if the time period is year-quarter
        * `"yw"` if the time period is year-week
    * `h` is an integer indicating the number of holdout/forecast periods
* NOTE 1: The results of this function should be saved to an object.  When doing so, the following objects are returned:
    * `$plot` contains the smoothing methods plot
    * `$acc` contains the accuracy measures
    * `$fcresplot` contains a plot of the holdout period residuals
    * `$fcresid` is a dataframe of the holdout period residuals (seldom used separately)
* Note 2: The model names are:
    * **Mov.Ave** for the moving average model
    * **Exp.Smooth** for the exponential smoothing model
    * **H-W.Add** for the Holt-Winters Additive model
    * **H-W.Mult** for Hold-Winters Multiplicative model

``` {r}
smooth <- smoothfc(msales, "t", "sales", "ym", 12)
smooth$plot
smooth$acc
smooth$fcresplot
```

## `linregfc` User Defined Function

* Usage: `linregfc(data, tvar="", obs="", datetype=c("ym", "yq", "yw"), h= )`
    * `data` is the name of the dataframe with the time variable and measure variable
    * `tvar` is the variable that identifies the time period (in quotes)
    * `obs` is the variable that identifies the measure (in quotes)
    * `datetype` can be one of three options:
        * `"ym"` if the time period is year-month
        * `"yq"` if the time period is year-quarter
        * `"yw"` if the time period is year-week
    * `h` is an integer indicating the number of holdout/forecast periods
* NOTE 1: The results of this function should be saved to an object.  When doing so, the following objects are returned:
    * `$plot` contains the regression methods plot
    * `$acc` contains the accuracy measures
    * `$fcresplot` contains a plot of the holdout period residuals
    * `$fcresid` is a dataframe of the holdout period residuals (seldom used separately)
* Note 2: The model names are:
    * **Lin.Reg.Trend** for the linear regression with trend model
    * **Line.Reg.Seas.Trend** for linear regression with trend and seasonality model
    
``` {r}
linreg <- linregfc(msales, "t", "sales", "ym", 12)
linreg$plot
linreg$acc
linreg$fcresplot
```

## `fccompare` User Defined Function

* Usage: `fccompare(results, models)`
    * `results` is a list of the stored methods results; create the list with the `list` function, such as:<br>`results <- list(naive, smooth, linreg)`
    * `models` is a vector of the models you want to compare; create the vector with the `c()` function, such as:<br>`models <- c("Naive", "Mov.Ave", "Lin.Reg.Trend")`
* NOTE 1: The function returns two items that will be directly displayed or can be saved to an object:  
    * `$acc` contains the accuracy measures
    * `$fcresplot` contains a plot of the holdout period residuals

``` {r}
results <- list(naive, smooth, linreg)
models <- c("Naive", "Mov.Ave", "Lin.Reg.Trend")
fccompare(results, models)
```